---
title: "League of Legends Data Wrangling"
author: "Simon Thornewill von Essen"
date: "Janurary 25, 2018"
output: html_document
---

# Introduction

In this R-markdown file, I am going to be cleaning and wrangling the data for
the League of Legends Ranked Matched Dataset. (See the references in the 
README.md file in this directory for the link to this dataset.)

First thing I am going to do is to set my working directory, libraries I will be
using and import the relevant data.

```{r Import Data and Packages, warning=FALSE, warning=FALSE, message=FALSE}
# Set Working Directory
setwd("~/Documents/git-repos/udacity/Udacity-Project---Create-a-Tableau-Story")

# Import Libraries
library(dplyr)
library(data.table)

# Import Data
df_champs <- read.csv("data/pre-wrangle/champs.csv")
df_matches <- read.csv("data/pre-wrangle/matches.csv")
df_participants <- read.csv("data/pre-wrangle/participants.csv")
df_stats1 <- read.csv("data/pre-wrangle/stats1.csv")
df_stats2 <- read.csv("data/pre-wrangle/stats2.csv")
df_teambans <- read.csv("data/pre-wrangle/teambans.csv")
df_teamstats <- read.csv("data/pre-wrangle/teamstats.csv")
```

# Getting to know the data

This dataset takes the form of a relational database to save on space.

I should keep and remove data based on what I am interested in.

First, we should take a look at the structure of these dataframes so that we
have some understanding of what sort of data they contain and what sorts of 
questions I would like to ask the data.

One drawback of this dataset is that it is not really clear what ID relates with
other dataframes. This is something the creator of this dataset could have been
more careful about. (Although I was able to piece things together in the end.)

## champs.csv

```{r}
str(df_champs)

```

We can see that this is a relatively small dataframe containing only the 
various champions in the game and their champion IDs.

I used to play a lot league of legends 5 or 6 years ago, so I am familliar with
most of the champions, their roles and their strengths and weaknesses. I will 
not be very familliar with newer characters or any changes to the meta while I
was away.

## matches.csv

```{r}
str(df_matches)

```

The matches file has a little more information inside of it. Containing id, 
gameid (matched with teambans and teamstats) in df_participants), platformid 
(server on which the game was played), queueid (only ranked-solo games shown in 
dataset), seasonid, duration (in seconds), creation (timestamp in $\mu$seconds) 
and game version (patch). 

Some of these columns are less interesting such as version, if I want to track 
the meta-game across seasons then seasonid will be important to keep. If I want
to only investigate the current meta-game then I should make sure I create 
appropriate filters for the data.

Columns such as "queueid" and "platformid" seem relatively useless 

## participants.csv

```{r}
str(df_participants)

```

Note that ss stands for "summoner spell" which is a spell that is selected 
during character selection and is independent of chosen champion. These are 
spells with relatively long cooldowns (a length of time where you cannot use 
this spell after casting it) and are relatively powerful. Two populat examples 
of such spells are finishing moves such as "ignite" and "flash" to help save a
champion in tight spots.

This dataframe contains a number of foreign keys such as "matchid", which relates
to teambans and teamstats, "championid" which corresponds with id in the champs 
dataframe and ID which corresponds to the df_stats dataframes.

role and position seem to have a slight overlap, the difference being that the
"BOT" position could refer to either being the "AD carry" or the "support" roles
denoted by "duo_carry" and "duo", respectively.

The most interesting columns in this set is the role and the position played by
each player and champion. 

## stats1.csv and stats2.csv

```{r}
str(df_stats1)

```

Stats is short for the player-stats, and contains general statistics of what 
each player did during a game.

What columns I am interested in from this dataset will depend on what question
I decide to answer. 

Different metrics will reveal different parts of the data.

## teambans.csv

```{r}
str(df_teambans)

```

Before a ranked match begins, a couple of champions are banned from selection.
This means that if a team wants to block a certain kind of strategy  or if there
is a champion that is really overpowered then they can make sure that they do 
not joint he fray until Riot Inc. corrects the cause of their overpowered 
nature.

## teamstats.csv

```{r}
str(df_teamstats)

```

It should be noted that teamstats and teambans are both related by matchid and
teamid. 

# What questions do I want to find out?

I want to know what the strongest and weakest characters in the meta are. 
However, it might not be easy to compare champions since a number of champions
can take multiple positions and so I should consider what the strongest 
characters for each position might be. 

It should be noted that there are a couple of common metrics which will decide
how strong a champion is, how much damage they do throughout the course of a 
game, how many kills or assists they have, how little they have died, how 
much/little damage they have taken and how many minion/creature kills they have.

This is actually a pretty good way to measure most roles except for the support
classes which are better guaged by how much damage they take, how much they have
healed themselves and other champions, how many assists (not kills) they have 
and how few deaths they have.

It is also worth noting that the characters that are most commonly banned will
also be champions that are incredibly powerful, so that should be investigated
as well. At the very least, there is no point in choosing a brokenly powerful
champion when it will get banned right away.

If I want to investigate the meta then I should make sure that changes make to 
the game that happen throughout patches which make certain champions more
powerful than each other are taken into account. Although this might not vary 
too much when considering individual patches, many large changes take place 
between seasons which means that I should make sure to take the latest season.


# Removing unecessary data

## df_matches

I will begun my cleaning at the matches stage since there are a number of 
columns that I would like to eliminate and because I can limit the season number
at this section and then use the IDs remaining to figure out what rows I should
remove from "Participants" and so on.

```{r}
# Filter Season
df_matches <- subset(df_matches, seasonid==8)

# Name Columns to remove
remove_cols <- c("platformid", "queueid", "creation", "version", "seasonid")

# Remove columns
df_matches <- df_matches[ , !(names(df_matches) %in% remove_cols)]

# Remove vector
rm(remove_cols)

# Rename gameid to matchid
df_matches <- rename(df_matches, matchid=id)

# Remove matchids that don't appear in df_participants
df_participants <- df_participants[df_participants$matchid %in% df_matches$matchid, ]

```

## df_participants

I also that the Jungle position has "none" in the none field. I just want to 
make sure that this is consistent.

We can see that the number of distinct roles for the "JUNGLE" position is 
`r n_distinct(subset(df_participants, position=="JUNGLE")$role)`, while the 
number of positions for the "NONE" role is
`r n_distinct(subset(df_participants, role=="NONE")$position)`.

This means that there are other positions that have the "NONE" role.

```{r}
count(subset(df_participants, role=="NONE"), position)

```

We can see here that there is a high probability that a player who has the 
"NONE" position has a high chance of being a jungler, but sometimes the lanes 
also have this position. My guess is that these values were determined by where
the players were located towards the start of the game and these players went
absent to attack players on other lanes.

Indeed, further investigation into the various roles that a position can have,
we can see that most positions end up having multiple roles, which seems to
complicate things a bit.

However, I will not wrangle the data here because, hopefully, the amount of 
non-standard plays are relatively low and so they should not interfere with the 
data too much. 

## df_teambans

Looking at the teambans dataframe, most of the data seems to be useful and so
I should keep it.

The only thing I want to make sure I do is to remove gameids from df_teambans
that are not found in df_participants

```{r}
# Remove matchids that don't appear in df_participants
df_teambans <- df_teambans[df_teambans$matchid %in% df_participants$matchid, ]

```

## df_teamstsats

The same goes for teamstats. I could remove the columns which include "harry"
(short for herald, apparently) because I never played a game where it was 
relevant and it makes the files a little better than they need to be

```{r}
# Name Columns to remove
remove_cols <- c("firstharry", "harrykills")

# Remove columns
df_teamstats <- df_teamstats[ , (!names(df_teamstats) %in% remove_cols)]

# Remove vector
rm(remove_cols)

# Remove matchids that don't appear in df_participants
df_teamstats <- df_teamstats[df_teamstats$matchid %in% df_participants$matchid, ]

```

## df_stats 1 and 2

Finally, I would like to remove a couple of columns from df_stats because it is
unlikely that I will use all of the columns and I can save on space in random
access memory this way as well.

```{r}
# Name Columns to remove
remove_cols <- c("item1", "item2", "item3", "item4", "item5", "item6", 
                 "trinket", "doublekills", "tripplekills", "quadrakills", 
                 "pentakills", "legendarykills", "visionscore", "pinksbought")

# Remove columns
df_stats1 <- df_stats1[ , (!names(df_stats1) %in% remove_cols)]
df_stats2 <- df_stats2[ , (!names(df_stats2) %in% remove_cols)]

# Remove vector
rm(remove_cols)

# Remove ids that don't appear in df_participants
df_stats1 <- df_stats1[df_stats1$id %in% df_participants$id, ]
df_stats2 <- df_stats2[df_stats2$id %in% df_participants$id, ]

# Concatinate dataframes like a madman
df_stats <- rbind(df_stats1, df_stats2)
rm(df_stats1, df_stats2)

```

I removed items because I am currently unware of the typical builds used by each
role. Investigating the different builds is outside of the scope of this 
investigation and so I removed them. The information for multi-kills is also
encoded in other columns such that they become a bit redundant. The same for
vision score and potion/drinks bought(?).

# Checking for missing values

```{r}
# Checking for missing values

## df_champs
sapply(df_champs, function(x) sum(is.na(x)))

## df_participants
sapply(df_participants, function(x) sum(is.na(x)))

## df_stats
sapply(df_stats, function(x) sum(is.na(x)))

## df_teambans
sapply(df_teambans, function(x) sum(is.na(x)))

## df_teamstats
sapply(df_teamstats, function(x) sum(is.na(x)))

```

Luckily, it seems like there are no missing values inside of this dataset, which
is very nice!

# Checking for duplicated values

```{r}
# Check for duplicates
sum(duplicated(df_champs))
sum(duplicated(df_participants))
sum(duplicated(df_stats))
sum(duplicated(df_teambans))
sum(duplicated(df_teamstats))

```

It also seems that there are no duplicated values, which is good.

Otherwise, it also seems that the dataframes in this dataset are consistent and
appropriate, which is also good. 

# Exporting and Saving Data

```{r}
write.csv(x=df_stats, file="data/wrangled/stats.csv")
write.csv(x=df_matches, file="data/wrangled/matches.csv")
write.csv(x=df_champs, file="data/wrangled/champs.csv")
write.csv(x=df_participants, file="data/wrangled/participants.csv")
write.csv(x=df_teambans, file="data/wrangled/teambans.csv")
write.csv(x=df_teamstats, file="data/wrangled/teamstats.csv")

```
